Pulsar detection - Machine learning and pattern recognition exam

### 2023-03-11 report/paper.tex
Completed report

\documentclass[12pt,a4paper]{article}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{float}
\usepackage{multirow}
\usepackage{subfig}
\usepackage{alphalph}
\renewcommand\thesubfigure{\alphalph{\value{subfigure}}}

\begin{document}
\title{Pulsar detection - Machine learning and pattern recognition exam}
\author{Alberto Baroso - s296520}
\date{2022-07-13}
\maketitle
\clearpage

\begin{abstract}
    The Pulsar dataset \cite{stw656} is a collection of 17.898 pulsar candidates of which 1.639 are real pulsars and 16,259 are spurious examples caused by RFI/noise.
    A pulsar is a neutron star that emits beams of electromagnetic radiation.
    As pulsars rotate, their emission beam sweeps across the sky, a periodically repeated pattern of broadband radio emission can be detected as this beam crosses our line of sight.
    Thus pulsar search involves looking for periodic radio signals with large radio telescopes.
    The goal of this project is to use machine learning techniques to solve the binary classification problem of detecting pulsars from the Pulsar dataset.
\end{abstract}
\clearpage

@@

\subsection{Features}

Pulsar candidates are described by a class label and 8 continuous features extracted from radio signals collected by radio telescopes. \vspace{0.2cm}

The first four are simple statistics obtained from the integrated pulse profile (folded profile).
This is an array of continuous features that describe a longitude-resolved version of the signal that has been averaged in both time and frequency (see \cite{Lyon2016} for more details).
The remaining four features are similarly obtained from the DM-SNR curve (again see \cite{Lyon2016} for more details).
\vspace{0.2cm}
The features are listed below:

\begin{enumerate}
    \item Mean of the integrated profile.
    \item  Standard deviation of the integrated profile.
    \item Excess kurtosis of the integrated profile.
    \item Skewness of the integrated profile.
    \item Mean of the DM-SNR curve.
    \item Standard deviation of the DM-SNR curve.
    \item Excess kurtosis of the DM-SNR curve.
    \item Skewness of the DM-SNR curve.
    \item Class
\end{enumerate}

The class labels used are 0 (non pulsar) and 1 (pulsar).

@@

An initial analysis of the raw features of the training data shows that:
\begin{itemize}
    \item Features have mainly irregular distributions
    \item Few features, the ones for non pulsars in figures (a), (b) and (g), present distributions similar to Gaussians
    \item Looking at the graph (b) it's possible to deduce that the "Standard Deviation of the integrated profile" could be one of the less informative features as the histograms of the two classes share one of the largest overlapping areas compared to the other graphs. But since it also has a good amount of non overlapping areas it can still provide useful information for classification purposes.
    \item Grphs (c) and (d) show that "Excess of kurtosis of integrated profile" and "Skewness of integrated profile" could be among the best features for sample discrimination, as they have little overlapping areas of the histograms.
    \item Some features, mainly the ones in graphs (b), (d) and (h), are characterized by the presence of significantly large outliers. Similar claims can be made by looking at the ranges of minimum and maximum values for the corresponding features
    \item We can expect classification algorithms to provide less than ideal results due to the presence of outliers.
\end{itemize}

% As a result, we further pre-process the data by "Gaussianizing" the features.

\subsection{Feature pairs}

@@

% 
% Classes are highly imbalanced.
%

Observing the feature pairs it's possible to note that there isn't one that allows to find a clear separation of the two classes. 
\vspace*{0.5cm}
The combination of features represented in graphs (n), (p), (ab), (af) and (ah) get close to identify regions of space where samples are well separated.
\vspace*{0.5cm}\\
The least useful pairs of features are represented in graphs (q), (s), (u) and (ac), where samples are almost entirely overlapping.

% TODO: gaussianize features
% TODO: feature correlation: compute pearson correlation heatmap

\clearpage
\bibliographystyle{unsrt}
\bibliography{citations}

\end{document}

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{float}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{subfig}
\usepackage{alphalph}
\usepackage{scrextend}
\usepackage{xcolor}
\renewcommand\thesubfigure{\alphalph{\value{subfigure}}}

\begin{document}
\title{Pulsar detection - Machine learning and pattern recognition exam}
\author{Alberto Baroso \\ s296520}
\vspace{1cm}
\date{2023-01-31}
\maketitle
\clearpage

\begin{abstract}
    The Pulsar dataset \cite{stw656} is a collection of 17.898 pulsar candidates of which 1.639 are real pulsars and 16,259 are spurious examples caused by RFI/noise.
    A pulsar is a neutron star that emits beams of electromagnetic radiation.
    As pulsars rotate, their emission beam sweeps across the sky, a periodically repeated pattern of broadband radio emission can be detected as this beam crosses our line of sight.
    Thus pulsar search involves looking for periodic radio signals with large radio telescopes.
    \\ \\
    The goal of this project is to find the best machine learning techniques to solve the binary classification problem of detecting pulsars from the Pulsar dataset.
    The analysis starts with the exploration of the dataset and its features, it proceeds with the application of dimensionality reduction and preprocessing techniques,
    different kinds of models are then evaluated and their scores are calibrated, finally, experiments are carried out on the test dataset to evaluate if the choices made during training were optimal.
\end{abstract}
\clearpage

@@

\subsection{Datasets}

The Pulsar dataset \cite{stw656} contains 17.898 pulsar candidates of which 1.639 are real pulsars and 16,259 are spurious examples.
The distribution of the two classes is highly unbalanced.
The data has been split into a training set and a test set:
\begin{itemize}
    \item The training set contains 8929 samples, of which 8108 are non-pulsars and 821 are pulsars.
    \item The test set contains 8969 samples, of which 8151 are non-pulsars and 818 are pulsars.
\end{itemize}
The class labels used are: 0 for non-pulsar samples and 1 for pulsars.

\subsection{Feature descriptions}

Pulsar candidates are described by 8 continuous features extracted from radio signals collected by radio telescopes.
The first four are simple statistics obtained from the integrated pulse profile (folded profile).
This is an array of continuous features that describe a longitude-resolved version of the signal that has been averaged in both time and frequency (see \cite{Lyon2016} for more details).
The remaining four features are similarly obtained from the DM-SNR curve (again see \cite{Lyon2016} for more details).
The features are listed below:

\begin{enumerate}
    \item Mean of the integrated profile.
    \item Standard deviation of the integrated profile.
    \item Excess kurtosis of the integrated profile.
    \item Skewness of the integrated profile.
    \item Mean of the DM-SNR curve.
    \item Standard deviation of the DM-SNR curve.
    \item Excess kurtosis of the DM-SNR curve.
    \item Skewness of the DM-SNR curve.
\end{enumerate}

@@

An initial analysis of the raw features of the training data shows that:
\begin{itemize}
    \item Features mainly follow irregular distributions
    \item Few features, the ones for non-pulsars in figures (a), (b), and (g), present distributions similar to Gaussians
    \item Looking at graph (b) it's possible to deduce that the "Standard Deviation of the integrated profile" could be one of the less informative features as the histograms of the two classes share one of the largest overlapping areas compared to the other graphs. But since it also has a good amount of non-overlapping areas it can still provide useful information for classification purposes.
    \item Graphs (c) and (d) show that "Excess of kurtosis of integrated profile" and "Skewness of integrated profile" could be among the best features for sample discrimination, as they have little overlapping areas of the histograms.
    \item Values for non-pulsars are generally concentrated around some values, while pulsars present a wider range of values.
    \item Some features, mainly the ones in graphs (b), (d), and (h), are characterized by the presence of significantly large outliers. Similar claims can be made by looking at the ranges of minimum and maximum values for the corresponding features. It's possible to expect classification algorithms to provide less than ideal results due to the presence of outliers.
\end{itemize}

\subsection{Feature correlation}

@@

Observing the feature pairs it's possible to note that there isn't one that allows finding a clear separation of the two classes. \\ \\
The combination of features represented in graphs (n), (p), (ab), (af), and (ah) get close to identifying regions of space where samples are well separated. \\ \\
The least useful pairs of features are represented in graphs (q), (s), (u), and (ac), where samples are almost entirely overlapping. \\ \\
The plots also show that some pairs of features share similar distributions, implying that there is some sort of correlation between them. \\ \\
Overall, it's reasonable to believe that there are enough pairs of features that separate classes well enough to obtain good accuracy for the trained models.

\clearpage

Pearson correlation \( \frac{Cov(X,Y)}{\sqrt{Var(X)}\sqrt{Var(Y)}} \) was used to explore the correlation between all pairs of features.
The absolute value of the Pearson correlation between all pairs of features is shown in the following heatmaps: \\

\begin{figure}[H]
    \caption{Pearson correlation heatmap - RAW features}
    \begin{center}
        \begin{tabular}{ccc}
            \hspace*{-65pt}
            \subfloat[Whole dataset]{\includegraphics[width = 160pt]{img/pearson_correlation/raw_heatmap.png}}  &
            \subfloat[Pulsars]{\includegraphics[width = 160pt]{img/pearson_correlation/raw_pulsar_heatmap.png}} &
            \subfloat[Non-pulsars]{\includegraphics[width = 160pt]{img/pearson_correlation/raw_non_pulsar_heatmap.png}} \\
        \end{tabular}
    \end{center}
\end{figure}

\begin{figure}[H]
    \caption{Pearson correlation heatmap - Z-normalized features}
    \begin{center}
        \begin{tabular}{ccc}
            \hspace*{-65pt}
            \subfloat[Whole dataset]{\includegraphics[width = 160pt]{img/pearson_correlation/z-normalization_heatmap.png}} &
            \subfloat[Pulsars]{\includegraphics[width = 160pt]{img/pearson_correlation/z-normalization_pulsar.png}}        &
            \subfloat[Non-pulsars]{\includegraphics[width = 160pt]{img/pearson_correlation/z-normalization_non_pulsar.png}}  \\
        \end{tabular}
    \end{center}
\end{figure}

Both raw features and z-normalized features heatmaps highlight the presence of strong correlation between some pairs of features. In particular between features:

\begin{enumerate}
    \item (0) Mean of the integrated profile - (2) Excess kurtosis of the integrated profile.
    \item (1) StDev of the integrated profile - (3) Skewness of the integrated profile.
    \item (4) Mean of the DM-SNR curve - (5) Standard deviation of the DM-SNR curve.
    \item (6) Excess kurtosis of the DM-SNR curve - (7) Skewness of the DM-SNR curve.
\end{enumerate}

In general, the first four features appear to be related to each other and this is because they are all measures of the integrated profile.
The same consideration applies to the last four features, while the first four and the last four present little correlation.
The presence of correlation between features suggests that some models, such as diagonal MVG classifiers, might benefit from the application of PCA to reduce the amount of features.
Applying PCA effectively reduces the overall feature correlation, below the heatmap obtained after the application of PCA to 4 dimensions is shown: \\

\begin{figure}[H]
    \caption{Pearson correlation heatmap - PCA features}
    \begin{center}
        \begin{tabular}{ccc}
            \hspace*{-65pt}
            \subfloat[Whole dataset]{\includegraphics[width = 160pt]{img/pearson_correlation/pca_heatmap.png}} &
            \subfloat[Pulsars]{\includegraphics[width = 160pt]{img/pearson_correlation/pca_pulsar.png}}        &
            \subfloat[Non-pulsars]{\includegraphics[width = 160pt]{img/pearson_correlation/pca_non_pulsar.png}}  \\
        \end{tabular}
    \end{center}
\end{figure}

\clearpage

\subsection{Applications}

Given no prior knowledge of the environment in which the final model will work,
3 different applications have been considered, one where the priors are balanced and two where the prior is biased towards one of the classes:
\begin{itemize}
    \item ( \(\tilde{\pi}\), Cfp, Cfn) = (0.5, 1, 1)
    \item ( \(\tilde{\pi}\), Cfp, Cfn) = (0.1, 1, 1)
    \item ( \(\tilde{\pi}\), Cfp, Cfn) = (0.9, 1, 1)
\end{itemize}

\section{Data preprocessing}

\subsection{K-fold cross-validation}

The K-fold approach has been used instead of the single-split validation technique to achieve more robust results, as the time required for K-fold is not excessive.
Models are compared by means of the Minimum Detection Cost function, a measure of the cost paid for making optimal decisions on the validation set. \\

The original training set has been split into 10 folds of similar size using the K-fold cross-validation technique.
At each iteration of the K-fold cross-validation algorithm, 9 folds have been employed for training and 1 for validation.
The data preprocessing techniques have been applied to the data after it was split into folds,
i.e. the transformation to be applied, e.g. PCA projection matrix, has been computed on the 9 train folds, then it has been applied to both training and validation folds.
This was done to avoid bias in the results, as splitting after preprocessing would have meant classifying data that the model had already seen.
\\ \\The procedure followed to train each model is the same:
\begin{enumerate}
    \item Divide training samples into K-folds, eventually applying preprocessing techniques (Computation performed only once)
    \item For each fold:
          \begin{itemize}
              \item Train a new model on the K-1 training folds
              \item Use the trained model to score the validation fold
          \end{itemize}
    \item Combine the scores of all folds
    \item Use the scores to compute minimum detection cost
\end{enumerate}

\subsection{Shuffling}

Before applying any preprocessing technique, the training dataset has been randomly shuffled to avoid any bias that could be introduced by the order of the samples.

\subsection{Z-Normalization}

All samples have been processed by means of Z-normalization, i.e. each feature has been transformed to have zero mean and unit variance.
$ x_i = \frac{x_i - \mu}{\sigma} $
$\mu$ and $\sigma$ are respectively the mean and the standard deviation of each feature describing the samples.

\subsection{Feature Gaussianization}

As an alternative preprocessing step to z-normalization, raw features have been "Gaussianized", i.e. their distributions have been transformed to be more similar to Gaussian distributions.
The Guasianization of a feature consists in computing a rank for each sample with respect to all samples and then computing the inverse of the cumulative distribution function of the standard normal distribution.
The rank of both training and evaluation data is computed with respect to the training data only.
\begin{figure}[H]
    \begin{center}
        \hspace*{-25pt}
        \begin{tabular}{ccc}
            \subfloat[]{\includegraphics[width = 200pt]{img/gaussianized_features/mean_of_ip}}                    &
            \subfloat[]{\includegraphics[width = 200pt]{img/gaussianized_features/stdev_of_ip.png}}                 \\
            \subfloat[]{\includegraphics[width = 200pt]{img/gaussianized_features/excess_kurtosis_of_ip.png}}     &
            \subfloat[]{\includegraphics[width = 200pt]{img/gaussianized_features/skewness_of_ip.png}}              \\
            \subfloat[]{\includegraphics[width = 200pt]{img/gaussianized_features/mean_of_dm-snr}}                &
            \subfloat[]{\includegraphics[width = 200pt]{img/gaussianized_features/stdev_of_dm-snr.png}}             \\
            \subfloat[]{\includegraphics[width = 200pt]{img/gaussianized_features/excess_kurtosis_of_dm-snr.png}} &
            \subfloat[]{\includegraphics[width = 200pt]{img/gaussianized_features/skewness_of_dm-snr.png}}          \\
        \end{tabular}
    \end{center}
\end{figure}
Gaussianization is computed over the whole training set, therefore class-conditional distributions might still not behave like a Gaussian distribution.
This effect is clearly visible in the plots above, non-pulsars follow more regular distributions (due to their higher number of samples), while pulsars show in most cases moderate deviations from the Gaussian assumptions.
Nonetheless, the new Gaussian-like distributions are likely to improve results of classifiers that assume data are Normal distributed.
This preprocessing step also helps reducing the negative effects of the outliers as they have been mapped as "closer" values on the tail-end of the Gaussian distributions.

\subsection{Linear Discriminant Analysis}
Linear Discriminant Analysis (LDA) is deemed not useful as a preprocessing step for this dataset as LDA is capable of finding at most c - 1 discriminant directions, where c is the number of classes.
Being this a binary classification problem, the maximum number of directions LDA can produce is 1.
Given the huge loss of information this would imply, results obtained on LDA-preprocessed data will likely be worse than the ones of other approaches.
Therefore, the use of LDA is not explored in this project.

\subsection{Principal Component Analysis}

The results from the correlation analysis suggest that some models might benefit from applying a dimensionality reduction technique such as PCA to obtain uncorrelated features.
Removing correlation between features could also improve model performances by helping to reduce overfitting.
Although, given the already small number of features, further reducing it might lead to a loss of information higher than the gain obtained by the removal of correlation.
The PCA projection matrix has been computed on the training dataset and applied to the training and evaluation datasets.

\clearpage

\begin{multicols}{2}
    {
        \noindent
        \begin{addmargin}[0em]{1em}
            The effects of different values of the hyperparameter m (number of components) for PCA have been evaluated by means of minimum DCF over different MVG models.
            Table 2 shows the results obtained with different values of m, types of Gaussian classifiers, and applications.
            Full diagonal models performances worsen as the number of components decreases but overall this deterioration is not excessive.
            Classifiers with diagonal covariance on the other hand slightly improve their performances as correlated features are removed, but their overall performance is worse than the full covariance models, meaning that the training dataset is large enough to reliably estimate the full covariance matrix and effectively capture the correlation in the data.
            Since full covariance models are capable of capturing these correlations it doesn't bring any advantage to reduce the dimensionality of the data. PCA is therefore not used in the following experiments.
        \end{addmargin}
    }
    {
        \noindent
        \renewcommand{\arraystretch}{1.2}
        \begin{tabular}{cccc}

            Model      & \(\tilde{\pi}\)=0.5  & \(\tilde{\pi}\)=0.1   & \(\tilde{\pi}\)=0.9   \\ [0.5ex]

            \hline
            \multicolumn{4}{c}{RAW Features - No PCA}                                         \\
            \hline
            Full Cov.  & 0.141                & 0.280                 & 0.680                 \\
            Diagonal   & 0.194                & 0.314                 & 0.723                 \\
            Tied Full  & {\color{red} 0.113 } & {\color{blue} 0.224 } & {\color{blue} 0.571 } \\
            Tied Diag. & 0.161                & 0.266                 & 0.591                 \\

            \hline
            \multicolumn{4}{c}{RAW Features - PCA (m = 7)}                                    \\
            \hline
            Full Cov.  & 0.174                & 0.302                 & 0.708                 \\
            Diagonal   & 0.220                & 0.628                 & 0.833                 \\
            Tied Full  & {\color{red} 0.153 } & {\color{blue} 0.280 } & {\color{blue} 0.612 } \\
            Tied Diag. & 0.172                & 0.306                 & 0.639                 \\

            \hline
            \multicolumn{4}{c}{RAW Features - PCA (m = 6)}                                    \\
            \hline
            Full Cov.  & 0.172                & 0.315                 & 0.662                 \\
            Diagonal   & 0.220                & 0.645                 & 0.818                 \\
            Tied Full  & {\color{red} 0.156 } & {\color{blue} 0.288 } & {\color{blue} 0.602 } \\
            Tied Diag. & 0.173                & 0.310                 & 0.618                 \\

            \hline
            \multicolumn{4}{c}{RAW Features - PCA (m = 5)}                                    \\
            \hline
            Full Cov.  & 0.183                & 0.423                 & 0.748                 \\
            Diagonal   & 0.203                & 0.607                 & 0.797                 \\
            Tied Full  & {\color{red} 0.179 } & {\color{blue} 0.315 } & {\color{blue} 0.592 } \\
            Tied Diag. & 0.187                & 0.345                 & 0.604                 \\

            \hline
            \multicolumn{4}{c}{RAW Features - PCA (m = 4)}                                    \\
            \hline
            Full Cov.  & 0.186                & 0.434                 & 0.683                 \\
            Diagonal   & 0.199                & 0.603                 & 0.799                 \\
            Tied Full  & {\color{red} 0.172 } & {\color{blue} 0.316 } & {\color{blue} 0.601 } \\
            Tied Diag. & 0.184                & 0.347                 & 0.611                 \\
        \end{tabular}
        \captionof{table}{MVG Classifiers - Tuning of PCA components}
    }
\end{multicols}

\section{Classification}

\subsection{Gaussian models}

\subsubsection{Multivariate Gaussian Classifier}
The MVG classifier uses full covariance matrices, i.e. a covariance matrix and a mean is estimated for each classes.
Given the large number of training samples, this kind of model is expected to provide good results compared to the other types of Gaussian classifiers.

\subsubsection{Tied Gaussian Classifier}
The tied MVG model uses the same covariance matrix for all classes.
This assumption shouldn't benefit the pulsar classification problem as the training set contains a good number of samples allowing covariance matrices to be reliably estimated for each class.

\subsubsection{Naive Bayes Gaussian Classifier}
The naive Bayes MVG model simplifies the estimation by assuming that the features are independent.
We thus expect the MVG classifier with naive Bayes assumption to perform worse that the previously seen
models because the correlation heatmaps highlighted a strong correlation between features.

\subsubsection{Tied Naive Bayes Gaussian Classifier}
The tied naive Bayes MVG combines the assumptions of feature independence and tied covariance matrices.
Given that the Naive assumption is not met, this model is also expected to have some among the worst performances.

\clearpage

\subsubsection{Results}

\begin{multicols}{2}
    {

        \begin{addmargin}[0em]{1em}
            As expected Diagonal models obtain the worst performances.
            Contrary to expectations though, tied covariance models provide better results than full covariance ones.
            This could be due to the fact that since the classes represent the same physical phenomena the covariance matrices of the two classes might be similar.
            Another possible explanation could be that linear separation rules, such as the one found by the tied model, are more appropriate than quadratic ones for this classification problem.

            Raw and Z-normalized features perform similarly.
            Gaussianized features improve the performances of the diagonal covariance model but overall perform worse than the Raw and Z-normalized features on the other models.

            \(\tilde{\pi}\) = 0.9 produces the worst results, \(\tilde{\pi}\) = 0.1 provides mediocre outputs but the best ones are obtained for application \(\tilde{\pi}\) = 0.5.
            The best performances are given by the Tied Full Covariance model on the Z-Normalized Features.
        \end{addmargin}
    }
    {
        \noindent
        \renewcommand{\arraystretch}{1.2}
        \begin{tabular}{@{}cccc@{}}
            Model      & \(\tilde{\pi}\)=0.5  & \(\tilde{\pi}\)=0.1   & \(\tilde{\pi}\)=0.9  \\ [0.5ex]

            \hline
            \multicolumn{4}{c}{RAW Features}                                                 \\
            \hline
            Full Cov.  & 0.141                & 0.280                 & 0.680                \\
            Diagonal   & 0.194                & 0.314                 & 0.723                \\
            Tied Full  & {\color{red} 0.113 } & {\color{red} 0.224 }  & 0.571                \\
            Tied Diag. & 0.161                & 0.267                 & 0.591                \\

            \hline
            \multicolumn{4}{c}{Z-Normalized Features}                                        \\
            \hline
            Full Cov.  & 0.141                & 0.280                 & 0.680                \\
            Diagonal   & 0.194                & 0.314                 & 0.723                \\
            Tied Full  & {\color{red} 0.113 } & {\color{red} 0.224 }  & 0.571                \\
            Tied Diag. & 0.161                & 0.267                 & 0.591                \\

            \hline
            \multicolumn{4}{c}{Gaussianized Features}                                        \\
            \hline
            Full Cov.  & 0.154                & 0.245                 & 0.717                \\
            Diagonal   & 0.152                & 0.277                 & 0.602                \\
            Tied Full  & 0.132                & {\color{blue} 0.237 } & {\color{red} 0.532 } \\
            Tied Diag. & 0.163                & 0.292                 & 0.622                \\
        \end{tabular}
        \captionof{table}{MVG Classifiers - Comparison of results}
    }
\end{multicols}

\subsection{Logistic Regression}

The logistic regression model is trained by minimizing the sum of the logistic loss over all samples.
The L-BFGS algorithm has been used as numerical solver to find the minimum.
The logistic regression model makes no assumptions on data distribution but requires a hypothesis
on the shape of the separation rule to find. A linear separation surface has been assumed: \[ w^T x+b \]
Where w is a vector perpendicular to the separation surface and b is a bias that tells where on w the separation rule lies.

To handle class unbalance, a prior-weighted regularized version of the objective function has been used: \\

\begingroup{
\setlength{\thinmuskip}{0mu}
\hspace*{-30pt}
$J(w, b)=\frac{\lambda}{2}\|w\|^2+\frac{\pi_T}{n_T} \sum\limits_{i=\left.1\right| c_i=1 }^n \log \left(1+e^{-z_i (w^Tx_i+b)}\right)+\frac{1-\pi_T}{n_F} \sum\limits_{i=\left.1\right| c_i=0 }^n \log \left(1+e^{-z_i (w^Tx_i+b)}\right)$ \\ \\
}\endgroup
\(\lambda\) is the regularization coefficient, \(\pi_T\) is the prior probability of the positive class, \(n_T\) is the number of positive samples, \(n_F\) is the number of negative samples, \(c_i\) is the class of the \(i\)-th sample.

\subsubsection{Hyperparameter tuning: \(\lambda\)}

Small values of \(\lambda\) lead to poor generalization, while higher values of \(\lambda\) allow finding small values of \(\|w\|\) but worse separation rules.
To find the best value of \(\lambda\) the minDCF has been plotted for different applications and values of \(\lambda\) on a logarithmic scale.
For the purpose of evaluating different values of \(\lambda\), only the main application was considered (\(\pi_T = \frac{1}{2}\))

\begin{figure}[H]
    \begin{center}
        \hspace*{-25pt}
        \begin{tabular}{ccc}
            \includegraphics[width = 200pt]{img/evaluation_plots/lambda-lr-z-normalized.png} &
            \includegraphics[width = 200pt]{img/evaluation_plots/lambda-lr-gaussianized.png}   \\
        \end{tabular}
    \end{center}
\end{figure}

The best results are obtained with \(\lambda = 0 \) for all the applications, and both Z-Normalized Gaussianized features.
The reason why regularization is not helpful can be found in the small number of dimensions, which implies little risk of overfitting.
Even if regularization provides little to no improvements, in the following experiments \(\lambda\) has been set to \(10^{-6}\).

\clearpage

\subsubsection{Results}

\begin{center}
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{@{}cccc@{}}
        Model                                                    & \(\tilde{\pi}\)=0.5   & \(\tilde{\pi}\)=0.1  & \(\tilde{\pi}\)=0.9   \\ [0.5ex]

        \hline
        \multicolumn{4}{c}{RAW Features}                                                                                                \\
        \hline
        Logisitc Regression \((\lambda = 10^{-6}, \pi_T = 0.5)\) & {\color{blue} 0.116 } & 0.219                & {\color{blue} 0.530 } \\
        Logisitc Regression \((\lambda = 10^{-6}, \pi_T = 0.1)\) & {\color{blue} 0.114 } & {\color{red} 0.213 } & 0.538                 \\
        Logisitc Regression \((\lambda = 10^{-6}, \pi_T = 0.9)\) & {\color{blue} 0.118 } & 0.218                & 0.534                 \\

        \hline
        \multicolumn{4}{c}{Z-normalized Features}                                                                                       \\
        \hline
        Logisitc Regression \((\lambda = 10^{-6}, \pi_T = 0.5)\) & {\color{blue} 0.116 } & 0.219                & {\color{blue} 0.530 } \\
        Logisitc Regression \((\lambda = 10^{-6}, \pi_T = 0.1)\) & {\color{red} 0.113 }  & {\color{red} 0.213 } & 0.537                 \\
        Logisitc Regression \((\lambda = 10^{-6}, \pi_T = 0.9)\) & 0.120                 & 0.220                & 0.537                 \\

        \hline
        \multicolumn{4}{c}{Gaussianized Features}                                                                                       \\
        \hline
        Logisitc Regression \((\lambda = 10^{-6}, \pi_T = 0.5)\) & 0.126                 & 0.238                & 0.534                 \\
        Logisitc Regression \((\lambda = 10^{-6}, \pi_T = 0.1)\) & 0.128                 & 0.229                & 0.537                 \\
        Logisitc Regression \((\lambda = 10^{-6}, \pi_T = 0.9)\) & 0.127                 & 0.238                & {\color{red} 0.524 }  \\

        \hline
        \multicolumn{4}{c}{Other models}                                                                                                \\
        \hline
        MVG Tied Full Covariance                                 & {\color{red} 0.113 }  & 0.224                & 0.571                 \\
    \end{tabular}
    \captionof{table}{Logistic Regression - Comparison of results}
\end{center}

Gaussianized features perform worse than Z-normalized and Raw features, as Logistic Regression doesn't require assumptions on the distribution of the data.
Table 4 shows that changing the value of \(\pi_T\) has little effect on the results.
Once again \(\tilde{\pi}\) = 0.9 performs worse than \(\tilde{\pi}\) = 0.1, which in turn has worse performances than \(\tilde{\pi}\) = 0.5.
Best results are obtained for Z-normalized features, with \(\pi_T = 0.1\) and \(\tilde{\pi} = 0.5\).
The improvement of using \(\pi_T = 0.1\) instead of \(\pi_T = 0.5\) is very small, but this can be explained by the unbalance in the dataset towards non-pulsar samples.
Tied MVG with full covariance has similar performances to the best logistic regression model.

\subsection{Support Vector Machines}

The SVM problem was solved using its dual formulation: \\

\begin{center}
    $ \max_{\alpha} \alpha^T1-\frac{1}{2}\alpha^TH\alpha $ subject to $ 0\leq\alpha_i\leq C_i, i=1,..,n $
\end{center}

An updated version of the constraints has also been considered to handle the unbalance between the classes.
Constraints have been changed to use costs proportional to the priors defined by the application:
$ C_T = C \cdot \frac{\pi_T}{\pi_T^{emp}} $
$ C_F = C \cdot \frac{\pi_F}{\pi_F^{emp}} $
where $ C_i = C_T $ for pulsars and $ C_i = C_F $ for non-pulsar samples.
$\pi_T^{emp}$ and $\pi_F^{emp}$ are the empirical priors obtained from the training set.

\subsubsection{Hyperparameter tuning: C}
C represents the tradeoff between having good accuracy on the training set and finding a large margin (better generalization).
To find the best value of C, K-fold cross-validation has been employed to plot the minDCF for different applications and different values of C on a logarithmic scale:

\begin{figure}[H]
    \begin{center}
        \hspace*{-25pt}
        \begin{tabular}{ccc}
            \includegraphics[width = 200pt]{img/evaluation_plots/C-svm-z-normalized-unbalanced.png} &
            \includegraphics[width = 200pt]{img/evaluation_plots/C-svm-z-normalized-balanced.png}     \\
        \end{tabular}
    \end{center}
\end{figure}
\begin{figure}[H]
    \begin{center}
        \hspace*{-25pt}
        \begin{tabular}{ccc}
            \includegraphics[width = 200pt]{img/evaluation_plots/C-svm-gaussianized-unbalanced.png} &
            \includegraphics[width = 200pt]{img/evaluation_plots/C-svm-gaussianized-balanced.png}     \\
        \end{tabular}
    \end{center}
\end{figure}

Low values of C show the models start to underfit, while high values of C clearly show signs of overfitting.
The best results are obtained with C $\approx 10^{-1}$ for both balanced ($\pi_T = 0.5$) and unbalanced applications, and for both z-normalized and gaussianized features.

\subsubsection{Results}

Minimum DCF has been computed for different values of \(\pi_T\) and \(\tilde{\pi}\) for both z-normalized and gaussianized features.
Where $ \pi_T $ is the prior probability for the true hypothesis with which the model is trained and $ \tilde{\pi} $ represents the effective prior in the context where the model will be used.

\begin{center}
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{@{}cccc@{}}
        Model                                        & \(\tilde{\pi}\)=0.5   & \(\tilde{\pi}\)=0.1  & \(\tilde{\pi}\)=0.9  \\ [0.5ex]

        \hline
        \multicolumn{4}{c}{RAW Features}                                                                                   \\
        \hline
        Linear SVM (C = $10^{-1}$, unbalanced)       & 0.629                 & 0.803                & 0.954                \\
        Linear SVM (C = $10^{-1}$, $\pi_T = 0.5$)    & 0.592                 & 0.972                & 0.936                \\
        Linear SVM (C = $10^{-1}$, $\pi_T = 0.1$)    & 0.714                 & 1.000                & 0.931                \\
        Linear SVM (C = $10^{-1}$, $\pi_T = 0.9$)    & 0.599                 & 0.847                & 0.939                \\

        \hline
        \multicolumn{4}{c}{Z-normalized Features}                                                                          \\
        \hline
        Linear SVM (C = $10^{-1}$, unbalanced)       & 0.117                 & 0.222                & 0.584                \\
        Linear SVM (C = $10^{-1}$, $\pi_T = 0.5$)    & {\color{blue} 0.113 } & 0.223                & 0.558                \\
        Linear SVM (C = $10^{-1}$, $\pi_T = 0.1$)    & {\color{red} 0.111 }  & {\color{red} 0.214 } & 0.570                \\
        Linear SVM (C = $10^{-1}$, $\pi_T = 0.9$)    & 0.128                 & 0.234                & 0.527                \\

        \hline
        \multicolumn{4}{c}{Gaussianized Features}                                                                          \\
        \hline
        Linear SVM (C = $10^{-1}$, unbalanced)       & 0.135                 & 0.239                & 0.531                \\
        Linear SVM (C = $10^{-1}$, $\pi_T = 0.5$)    & 0.120                 & 0.251                & {\color{red} 0.519 } \\
        Linear SVM (C = $10^{-1}$, $\pi_T = 0.1$)    & 0.133                 & 0.237                & 0.531                \\
        Linear SVM (C = $10^{-1}$, $\pi_T = 0.9$)    & 0.130                 & 0.244                & 0.525                \\

        \hline
        \multicolumn{4}{c}{Other models}                                                                                   \\
        \hline
        MVG Tied Full Cov.                           & 0.113                 & 0.224                & 0.571                \\
        Log Reg \((\lambda = 10^{-6}, \pi_T = 0.1)\) & 0.113                 & 0.213                & 0.537                \\
    \end{tabular}
    \captionof{table}{Linear SVM - Comparison of results}
\end{center}

Table 5 shows that balancing classes has little impact on the outcome.
Raw features are the worst-performing ones, while z-normalized features perform better than gaussianized ones.
Best results are obtained for the z-normalized features with $\pi_T = 0.1$, \(\tilde{\pi}\) = 0.5, these are very close to the ones obtained for the best MVG and Logistic Regression models.

\subsubsection{Support Vector Machines with Polynomial Kernel}

The polynomial kernel $K(x_i, x_j) = (x_i^Tx_j + c)^d$ was used to train a linear separation surface in the expanded feature space, which corresponds to a quadratic separation rule in the original feature space.
Hyperparameters C and c were jointly optimized using a grid search, d was set to 2.

\begin{figure}[H]
    \begin{center}
        \hspace*{-25pt}
        \begin{tabular}{ccc}
            \includegraphics[width = 200pt]{img/evaluation_plots/polynomial-svm-z-normalized-C-c.png} &
            \includegraphics[width = 200pt]{img/evaluation_plots/polynomial-svm-gaussianized-C-c.png}   \\
        \end{tabular}
    \end{center}
\end{figure}
\vspace{-20pt}
\noindent
The polynomial kernel yields the best results when used with C=1 and c=0.1 for the z-normalized features and C=1, c=1 for the gaussianized features. \\

\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}cccc@{}}
    Model                                      & \(\tilde{\pi}\)=0.5  & \(\tilde{\pi}\)=0.1  & \(\tilde{\pi}\)=0.9  \\ [0.5ex]

    \hline
    \multicolumn{4}{c}{Z-normalized Features}                                                                       \\
    \hline
    Polynomial Kernel (c = 0.1, Unbalanced)    & 0.133                & 0.246                & 0.666                \\
    Polynomial Kernel (c = 0.1, $\pi_T = 0.5$) & {\color{red} 0.103 } & {\color{red} 0.220 } & {\color{red} 0.525 } \\
    Polynomial Kernel (c = 0.1, $\pi_T = 0.1$) & 0.129                & 0.228                & 0.662                \\
    Polynomial Kernel (c = 0.1, $\pi_T = 0.9$) & 0.141                & 0.297                & 0.594                \\
